---
title: "07_exercises"
format: pdf
editor: visual
---

## 07 • Exercises

Here are some exercises for the `data.table` package.

## → Exercise 01

Load the `data.table` package. Furtermore (install and) load the `AER` package and run the command `data("Fertility")`, which loads the dataset `Fertility` to your workspace. Turn it into a `data.table` object.

```{r}
library(data.table)
library(AER)
data("Fertility")

fert <- setDT(Fertility)
fert
```

## → Exercise 02

Select rows `35` to `50` and print to console its `age` and `work` entry.

```{r}
colnames(fert)

fert02 <- fert[35:50][, .(age, work)]
fert02
```

## → Exercise 03

Select the last row in the dataset and print to console.

```{r}
fert03 <- fert[nrow(fert)]
fert03
```

## → Exercise 04

Count how many women proceeded to have a third child.

```{r}
fert04 <- fert[morekids == "yes"]
fert04
```

## → Exercise 05

There are four possible gender combinations for the first two children. Which is the most common? Use the `by` argument.

```{r}
fert05 <- fert[, .(total = .N), by=c("gender1", "gender2")]
fert05[order(fert05)]
# It's clearly male-male, because it has the higher total amount
```

## → Exercise 06

By racial composition, what is the proportion of woman working four weeks or less in `1979`?

```{r}
#               
fert06 <- fert[, mean(work >= 4), by=c("afam", "hispanic", "other")]
fert06
```

## → Exercise 07

Use `%between%` to get a subset of woman between the age of 22 and 24 and calculate the proportion who had a boy as their firstborn.
